---
title: "mic_power-calc"
author: "Daniel Spakowicz"
date: "3/18/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(FDRsampsize)
library(ggplot2)
library(magrittr)
library(tidyr)
library(dplyr)
```

```{r sample size}
# Sample size
n.total <- 60

# Number of non-responders estimated at 25% of the total
est.response.rate <- 0.75

# Define the number in the smaller of the two arms
n.nr <- n.total * (1 - est.response.rate)

# Compliance rate
compliance <- 0.8
```

Correct for false positives using the Bonferroni method: 

> Bland JM, Altman DG. Multiple significance tests: the Bonferroni method. BMJ. 1995 Jan 21;310(6973):170. PMID: 7833759


```{r false positive rate}
# Define the false positive rate
sig.level <- 0.05
num.tests <- 100
typeIerror <- sig.level / num.tests
```

Define a range of fold changes and coefficients of variation consistent with other microbiome studies. 

```{r define range}
# Define ranges of fold changes and coefficients of variation 
# to calculate the power
fold.change <- c(2.1, 3, 4.2)
coef.variation <- c(0.4, 0.6, 0.8)
```

Calculate power using the method of Hart: 

> Hart SN, Therneau TM, Zhang Y, Poland GA, Kocher J-P. Calculating sample size estimates for RNA sequencing data. J Comput Biol. 2013 Dec;20(12):970–978. PMCID: PMC3842884


```{r calculate power}
power.l <- list()
for (c in 1:length(coef.variation)) {
  power.l[[as.character(coef.variation[c])]] <- 
    power.hart(n = n.nr * compliance, 
               alpha = typeIerror, 
               log.fc = log(fold.change),
               mu = rep(100, length(fold.change)), 
               sig = rep(coef.variation[c], length(fold.change)))
}

# Reformat the list for plotting
power.df <- bind_rows(power.l)
power.df$fold.change <- fold.change
```

Plot as a heatmap.

```{r plot}
# Plot as heatmap
power.df %>%
  gather(CV, Power, -fold.change) %>%
  ggplot(aes(x = factor(fold.change), y = CV)) +
  geom_tile(aes(fill = Power)) +
  scale_fill_gradient(low = "white", high = "blue") +
  geom_label(aes(factor(fold.change), CV, label = round(Power, 2))) +
  labs(x = "Fold Change",
       y = "Coefficient of Variation") +
  ggsave("figures/mic-power_cv-fc.png", height = 4, width = 4)
```

Text added to the protocol.

> Assuming 80% compliance rate for stool sample collection with 75% treatment response rate, to correct 100 tests, n=60 patients achieves 2.1 to 4.2-fold change (FC) between responders and nonresponders with coefficient of variation (CV) as 0.4 to 0.8 (Figure 1) by negative binomial Wald test \cite{https://pubmed.ncbi.nlm.nih.gov/23961961/}. This CV and FC range is consistent with the inter-individual variability observed in cross-sectional data in other microbiome studies \cite{https://pubmed.ncbi.nlm.nih.gov/31142858/}; the lower end of the range is likely in paired-sample analyses \cite{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6131952/}. We expect to have greater power to detect the difference in buccal swab samples due to the higher compliance rate. All calculations were made in R \cite{https://www.R-project.org/} using the package FDRsampsize \cite{https://CRAN.R-project.org/package=FDRsampsize}. Code related to this protocol is available at https://github.com/spakowiczlab/oromic. 